---
title: "Nick's R Cheat Sheet"
author: "Nicholas Vasile"
date: "`r Sys.Date()`"
output: 
  blogdown::html_page:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(magrittr) #for pipe
```
[PDF of many of Rstudio's cheat sheets](http://info.rstudio.com/G5X0Nw20UEEoN0YS0000JCn)

[Another R cheat sheet I found useful](https://github.com/startupjing/Tech_Notes/blob/master/R/R_language.md)

### Selecting attributes from a data frame
```{r}
# data [row, attribute_name]
iris[ 1, "Species"] 

# approach 1: use [[ form of extract operator to extract a column
iris[["Species"]] %>% 
  head()

# approach 2: use variable name in column dimension of data frame
iris[,"Species"] %>% 
  head()
# approach 3: use the $ form of extract operator. Note that since this
iris$Species %>% 
  head()

#Use individual variables in a pipe
ggplot2::diamonds %>% 
  .$cut %>% 
  head()

ggplot2::diamonds %>% 
  dplyr::pull(cut) %>% 
  head()
```

### subsetting rows
```{r}
# approach 1: use exact row references
# dataframe[rows, columns]
mtcars[20:22,]

# approach 2: use logic in the row dimension of reference
head(mtcars[mtcars$cyl == 4 & mtcars$am == 1,])
head(mtcars[mtcars[,"cyl"] == 4,])

# approach 3: use which() function
theSubsetRows <- which(mtcars$cyl == 4)
head(mtcars[theSubsetRows,])

# approach 4: use output from a function that returns a logical
#             array instead of row numbers as in the prior example
head(mtcars[!is.na(mtcars[,"cyl"]),])

# subset all rows with select  columns
# dataframe[, c("column1", "column2")]
head(mtcars[, c("mpg", "carb")])

# subset all rows where attribute matches
# dataframe[dataframe$attribute =="value", ] 
mtcars[mtcars$cyl==4,]


# select all rows and all columns except for a few others
# in this example remove mpg and disp, the first and third columns
mtcars[,-c(1,3)]

#alternate - if all rows you want to remove are next to each other
mtcars[,-(1:3)]

```

### Filtering / extracting / subsetting data frames based on attribute value
```{r}
mtcars[which(mtcars$cyl == 4), "mpg"]
subset(mtcars$mpg, mtcars$cyl == 4)
```


### Filtering for rows where at least one column is missing

```{r}
dplyr::starwars %>% 
  dplyr::filter(dplyr::if_any(dplyr::everything(), ~is.na(.x)))
```

### Filtering for rows where all columns have values

```{r}
dplyr::starwars %>% 
  tidyr::drop_na()
```


### Exploring data
```{r}
#summary stats from pastecs package
pastecs::stat.desc(mtcars)

#Counting NULL values in column
sum(is.na(mtcars$cyl))
```

### Prevent garbage characters when using read.csv on data exported from SQL
```{r, eval = FALSE}
#SOURCE: http://stackoverflow.com/questions/24568056/rs-read-csv-prepending-1st-column-name-with-junk-text
read.csv(file = "my_file.csv", fileEncoding = "UTF-8-BOM")
```

### List functions exported by a loaded package
```{r}
head(ls("package:base"))
```

### Count occurrences of unique values
```{r}
table(mtcars$cyl)
dplyr::count(mtcars, cyl)
as.data.frame(table(mtcars$cyl))

dplyr::count(mtcars, cyl, gear)
xtabs(~cyl + gear, mtcars)

#Get table with row and column sums
addmargins(table(mtcars$cyl,useNA = "always"))
#Get percentage table
addmargins(sort(prop.table(table(mtcars$cyl,useNA = "always"))))
```

### Check for duplicate values
```{r, eval = FALSE}
length(unique(nps$email))==nrow(nps)

#pull duplicate values
nycflights13::planes %>%
        dplyr::count(tailnum) %>%
        dplyr::filter(n > 1)

nycflights13::weather %>%
        dplyr::count(year, month, day, hour, origin) %>%
        filter(n > 1)
```

### Build and install vignettes
```{r, eval = FALSE}
devtools::install(build_vignettes = TRUE)
```

### String filtering
```{r}
#filter output of names - or any other character vector
row.names(mtcars)[grepl("Merc",row.names(mtcars))]
names(mtcars)[grepl("c", names(mtcars))]
grep("C", names(mtcars), ignore.case = TRUE, value = TRUE)
grep("C", names(mtcars), ignore.case = TRUE, value = FALSE)

#Filter with dplyr where row cointains string
dplyr::filter(ggplot2::diamonds, grepl('Good', cut))

#Filter column/attribute of data frame for regex text pattern match
nycflights13::airports %>%
  dplyr::filter(stringr::str_detect(name, "County")) %>% 
  head()

#Search for, and extract matches
pattern <- "[[:upper:]]"
stringr::sentences %>%
        stringr::str_subset(pattern) %>%
        stringr::str_extract_all(pattern) %>% 
  head()

#filter column/attriubte names based on string
ggplot2::diamonds %>%
  dplyr::select_at(dplyr::vars(dplyr::contains('c')))

#filter where attribute value starts with a string
mtcars %>% 
    tibble::rownames_to_column("model") %>% 
    dplyr::filter(stringr::str_detect(model, "^Merc"))

#rename columns/attributes if they contain certain pattern
#in this case replace '.' with '_'
#SOU
nycflights13::weather %>% 
  dplyr::rename_at(
    dplyr::vars(dplyr::contains('_')), dplyr::funs(stringr::str_replace(., "_", "\\."))
    ) %>% 
  names()
```

### String filtering when using a database and dbplyr
The previous mentions won't always work when  filtering against data in a database.
Reference: https://github.com/tidyverse/dplyr/issues/3090 

Solution from: https://stackoverflow.com/questions/38962585/pass-sql-functions-in-dplyr-filter-function-on-database/47198795#47198795
```{r}
#load some data into a temp database
con <- DBI::dbConnect(RSQLite::SQLite(), path = ":memory:")
DBI::dbWriteTable(con, "airports", nycflights13::airports)

#query the data - name contains "Island" - this is case sensitive
dplyr::tbl(con, "airports") %>% 
  dplyr::filter(name %like% "%Island%") %>% 
  head(5)
```

This works when I test against a Microsoft SQL database but not SQLite. Returns error:
`Error in stri_detect_regex(string, pattern, opts_regex = opts(pattern)) : object 'name' not found`

```{r, eval = FALSE}
#load some data into a temp database
con <- DBI::dbConnect(RSQLite::SQLite(), path = ":memory:")
DBI::dbWriteTable(con, "airports", nycflights13::airports)

#query the data - name contains "Island" - this is case sensitive
dplyr::tbl(con, "airports") %>% 
  dplyr::filter(stringr::str_detect(name, "Island")) %>% 
  head(5)
```

### Create permutations of items from multple vectors
```{r}
expand.grid(
    c(1, 2),
    c(3, 4),
    c(2, 3))
```
### Get class names of items in a dataframe/vector
```{r}
sapply(mtcars, class)
```
### Printing more than default rows/columns from a table
```{r}
mtcars %>% dplyr::as_tibble() %>% print(n=15, width = Inf)
```


### Search all objects, including functions, in global environment for string
```{r}
apropos("cars")
```

### Search key words or phrases in help pages, vignettes or task views
```{r, eval=FALSE}
RSiteSearch("Microsoft AND SQL")
```


### Update packages after updating R
```{r, eval = FALSE}
update.packages(ask = FALSE, checkBuilt = TRUE)
```

### Timezone stuff
```{r}
#Get timezone
Sys.timezone()

#Get all timezones
OlsonNames() %>% head()
```

### Calculate minutes between two times

```{r}
library(lubridate)
ymd_hms("2018-10-25 12:00:00") %--% ymd_hms("2018-10-25 13:00:00") / dminutes(1)
```

### filter for records within last n years

works with dbplyr too
this is not a great example but I wanted it to work for at least a few more years.

```{r}
hundred_years_ago <- lubridate::today() - lubridate::years(100)

nycflights13::flights %>% 
  dplyr::filter(time_hour >= hundred_years_ago)
```


### View data exported by a package
```{r, eval = FALSE}
data(package = "ggplot2")
```

### Browse vignettes for a given package
```{r, eval = FALSE}
browseVignettes(package = "dplyr")
```

### Open function documentation from w/in RStudio
Use F2

### Create a dataframe with random data
```{r}
#the call to set.seed ensures anyone who runs the code will get the same random number stream. 
set.seed(1)
data.frame(x = rnorm(4), y = rnorm(4), z = sample(LETTERS, 4))
```

### Details about built-in data sets
```{r, eval = FALSE}
library(help = "datasets")
data()
```

### Use Github to search for packages using a particular function
Put this into the GitHub search box to see how packages on CRAN use the llply() function from plyr
```
 "llply" user:cran language:R
```

### Conditional mutate
NOTE: should probably also look at recode here
```{r}
nycflights13::airlines %>%
    dplyr::mutate(
        name = dplyr::case_when(
            name == "Virgin America" ~ "Alaska Airlines Inc.",
            TRUE ~ name)
    )
```

### extracting nested list into a tibble
SOURCE: https://cfss.uchicago.edu/webdata004_simplifying_lists.html 
```{r}
my_list <- list(
    list(first = "nick", last = "vasile", weight = 170),
    list(first = "bob", last = "smith", weight = 150)
)

my_list

purrr::map_df(my_list, magrittr::extract)
```

### Scatterplot of all pairs
```{r}
pairs(mtcars)
```

### Interactively explore plots
```{r,eval=FALSE}
identify(mtcars$hp, mtcars$mpg, mtcars$mpg)
```

### Sort bar plot by counts
```{r}
nycflights13::flights %>% 
    ggplot2::ggplot(ggplot2::aes(x=forcats::fct_infreq(carrier))) +
    ggplot2::geom_bar()
```


```{r}
nycflights13::flights %>%
  ggplot2::ggplot(ggplot2::aes(x=forcats::fct_rev(
    forcats::fct_infreq(carrier)))) +
  ggplot2::geom_bar() +
  ggplot2::coord_flip()
```


### Sort bar plot by counts - when using stat = "identity"

```{r}
avg_delay_by_carrier <- nycflights13::flights %>% 
  dplyr::group_by(carrier) %>% 
  dplyr::summarize(mean_delay = mean(dep_delay, na.rm = TRUE))

avg_delay_by_carrier %>% 
  ggplot2::ggplot(ggplot2::aes(x=reorder(carrier, -mean_delay),
                               y = mean_delay)) +
  ggplot2::geom_bar(stat="identity") +
  ggplot2::labs(x = "carrier") # if you want to set x-axis to original attribute
```

```{r}
avg_delay_by_carrier <- 
  nycflights13::flights %>% 
  dplyr::group_by(carrier) %>% 
  dplyr::summarize(mean_delay = mean(dep_delay, na.rm = TRUE))

avg_delay_by_carrier %>% 
  ggplot2::ggplot(ggplot2::aes(x=reorder(carrier, mean_delay),
                               y = mean_delay)) +
  ggplot2::geom_bar(stat="identity") +
  ggplot2::coord_flip()
```

### Sort bar plots by counts, within facets, when using stat - "identity"

SOURCE: https://www.programmingwithr.com/how-to-reorder-arrange-bars-with-in-each-facet-of-ggplot/ 

```{r}
iris_gathered <- 
  tidyr::gather(iris, metric, value, -Species)

iris_gathered %>% 
  ggplot2::ggplot(ggplot2::aes(tidytext::reorder_within(Species, value, metric),
                      value)) +
  ggplot2::geom_bar(stat = 'identity') +
  tidytext::scale_x_reordered() +
  ggplot2::coord_flip() +
  ggplot2::facet_wrap(~ metric, scales = "free") +
  ggplot2::xlab("species")
```

### Color coded correlation table
```{r}
M <- cor(mtcars)
corrplot::corrplot(M, method="color")
```

with clustering

```{r}
corrplot::corrplot(M, order = "hclust")
```

### filter correlations at a cutoff value

```{r}
caret::findCorrelation(M, cutoff = 0.2, verbose = TRUE, names = TRUE)
```

I dont really like the formatting here - too hard to match columns - let's find a better way.

There is `corrr::correlations()` but not available on CRAN and there is also an open issue with the newest version of dplyr.

Using deprecrated `reshape2` package because tidyr doesn't handle matrixs. This is fine for now.

```{r}
reshape2::melt(M) %>%
  dplyr::filter(abs(value) > 0.2, Var1 != Var2) %>%
  dplyr::arrange(desc(abs(value))) %>%
  head()
```

### plot distribution of all variables

#### Just the numeric variables

SOURCE: (https://drsimonj.svbtle.com/quick-plot-of-all-variables)

```{r}
mtcars %>%
    purrr::keep(is.numeric) %>% 
    tidyr::gather() %>% 
    ggplot2::ggplot(ggplot2::aes(value)) + 
    ggplot2::facet_wrap(~key, scales = "free") + 
    ggplot2::geom_histogram()
```

#### all variables
```{r}
nycflights13::planes[,c(4,9)] %>%
  tidyr::gather() %>% 
  ggplot2::ggplot(ggplot2::aes(value)) + 
  ggplot2::facet_wrap(~key, scales = "free") + 
  ggplot2::geom_bar(stat = "count") + 
  ggplot2::coord_flip ()
```


### RMarkdown: insert date document was knitted

add this to the header: 

SOURCE: (https://stackoverflow.com/questions/23449319/yaml-current-date-in-rmarkdown)

### Plot percentage of attributes that are NA for each outcome

```{r}
data(Soybean, package = "mlbench")

# the first column is the outcome variable so we remove it
na_rates <- rowMeans(is.na(Soybean[,-1]))

soybean <- Soybean %>% 
  tibble::add_column(na_rate = na_rates)

soybean %>% 
    dplyr::group_by(Class) %>% 
    dplyr::summarize(mean_na_rate = mean(na_rate)) %>% 
    ggplot2::ggplot(ggplot2::aes(x=Class, y = mean_na_rate, fill = Class)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::coord_flip() +
    ggplot2::theme(legend.position = "none")
```

### Plot the pecentage of rows that has at least 1 NA attribute, by outcome

```{r}
data(Soybean, package = "mlbench")

na_rates <- rowMeans(is.na(Soybean[,-1]))

soybean <- Soybean %>%
    tibble::add_column(na_rates = na_rates) %>% 
    dplyr::mutate(has_nas = na_rates != 0 )

soybean %>% 
    dplyr::group_by(Class) %>% 
    dplyr::summarize(mean_has_nas = mean(has_nas)) %>% 
    ggplot2::ggplot(ggplot2::aes(x=Class, y = mean_has_nas, fill = Class)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::coord_flip() +
    ggplot2::theme(legend.position = "none")
```

## Plot the attributes (predictors) that are most likely to be missing

```{r}
data(Soybean, package = "mlbench")

predictor_na_rate <- data.frame(colMeans(is.na(Soybean[,-1]))) %>% 
    tibble::rownames_to_column(var = "predictor") %>% 
    dplyr::rename( na_rate = 2)

predictor_na_rate %>% ggplot2::ggplot(ggplot2::aes(x = predictor, y = na_rate,
                                                   fill = predictor )) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::coord_flip() +
  ggplot2::theme(legend.position = "none")
```

## Plot the attributes (predictors) that are most likely to be missing, by outcome

```{r}
data(Soybean, package = "mlbench")

soybean_l <- Soybean %>% 
  tidyr::gather(-Class, key = "predictor",value = "value")

soybean_l %>% dplyr::group_by(Class, predictor) %>% 
  dplyr::summarize(na_rate = mean(is.na(value))) %>% 
  ggplot2::ggplot(ggplot2::aes(x=predictor, y = na_rate, fill = predictor)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::coord_flip() +
  ggplot2::facet_wrap(~Class, scales = "free") +
  ggplot2::theme(legend.position = "none")
```

## Create a matrix that shows whether or not a particular combination of values is in the data
SOURCE: https://stackoverflow.com/a/37897416 
```{r}
mtcars %>% 
  tibble::rownames_to_column("name") %>% 
  dplyr::distinct(cyl, gear, has_gear = !is.na(name)) %>% 
  tidyr::complete( cyl, gear) %>% 
  dplyr::mutate(has_gear = dplyr::if_else(is.na(has_gear),FALSE, TRUE)) %>% 
  ggplot2::ggplot(ggplot2::aes(x=cyl, y =gear )) +
  ggplot2::geom_tile(ggplot2::aes(fill = has_gear),color = "white")
```

## Convert unix style epoch time to human readable time

```{r}
epoch_time <- 1505249866329

lubridate::as_datetime(epoch_time/1000)
```

## Clean Data: remove columns where no rows contain a value

```{r}

df <-data.frame(col_1 = c(NA, 1, 2, 3),
                col_2 = c(NA, NA, NA, NA),
                col_3 = c(1, 2, 3, 4)) 
  
df
  
case_missing_rate <- data.frame(colMeans(is.na(df))) %>%
        tibble::rownames_to_column(var = "column") %>%
        dplyr::rename(missing_rate = 2)

    all_missing_cols <- case_missing_rate %>%
        dplyr::filter(missing_rate == 1.0) %>%
        dplyr::pull(column)

    df %>%
        dplyr::select(-all_missing_cols)
```


## Print one plot for each data frame in a list column

There is probably a better way to meet the need.

My goals was to only show the values that are present in a given group.

`ggplot2::facet_wrap()` shows all values present in any group in the plots for 
all groups.

Step 1: make the list
```{r}
n_mtcars <- 
  mtcars %>%
  dplyr::group_by(cyl) %>% 
  tidyr::nest()

n_mtcars
```

Step 2: give the list members names
```{r}
names(n_mtcars$data) <- stringr::str_c(n_mtcars$cyl,"_cyl")
str(n_mtcars$data)
```

Step 3: Create plotting function

```{r}
plot_issues_by_epic <- function(df, item_name) {
  df %>% 
    ggplot2::ggplot(ggplot2::aes(gear)) +
    ggplot2::geom_bar() +
    ggplot2::coord_flip() +
    ggplot2::labs(title= item_name)
}
```

Step 4: Use your plotting function to create the plots

```{r}
purrr::map2(n_mtcars$data, names(n_mtcars$data),
     plot_issues_by_epic)
```

## One plot for each attribute in a data frame - different scale for each attribute

```{r}
datasets::airquality %>% 
  dplyr::filter(Month == 5) %>% 
  dplyr::select(-Month) %>% 
  tidyr::pivot_longer(-Day, names_to = "attribute", values_to = "value") %>% 
    ggplot2::ggplot(ggplot2::aes(Day, value)) +
    ggplot2::geom_line() +
    ggplot2::facet_grid(attribute ~ ., scales = "free")
```

## Get detailed information for each column in a database table

```{r, eval = FALSE}
odbc::odbcConnectionColumns(con, "table_name")
```

## Get just the columns and datatypes for each column in a database table

```{r, eval = FALSE}
odbc::odbcListColumns(con, table = "table_name")
```

## Get just the column names

```{r,eval = FALSE}
DBI::dbListFields(con, "table_name")
```

## Table of contents for rmarkdown document

Add this to the document header.

```{r, eval = FALSE}
output: 
    html_document:
        toc: true
        toc_depth: 6
        toc_float: true
```

## Get unique values from all non-numeric columns
```{r}
dplyr::starwars %>% 
  dplyr::select(!where(is.numeric)) %>% 
  purrr::map(unique)
```

