---
title: "Cheat Sheet: three ways to use dplyr to retrieve data from a Microsoft SQL
  Database"
author: 'Nick Vasile'
date: '2020-05-04'
slug: different-ways-to-use-dplyr-and-dbplyr-to-retrieve-data-from-a-microsoft-sql-database
categories: []
tags: []
---

## Motivation

I use R to extract data held in Microsoft SQL Server databases on a daily basis. 

When I first started I was confused by all the different ways to accomplish this task. I was a bit overwhelmed trying to choose the, "best," option given the specific job at hand.

I want to share what approaches I've landed on to help others who may want a simple list of options to get started with.

## Scope 

This post is about reading data from a database, not writing to one.

I prefer to use packages in the tidyverse so I'll focus on those packages.

While it's possible to generalize many of the concepts I write about here to other DBMS systems I will focus exclusively on Microsoft SQL Server. I hope this will provide simple, prescriptive guidance for those working in a similar configuration.

The data for these examples is stored using Microsoft SQL Server Express. Free download available [here](https://www.microsoft.com/en-us/sql-server/sql-server-downloads).

One last thing - these are a few options I populated my toolbox with. They have served me well over the past two years as an analyst in an enterprise environment, but are definitely not the only options available.

## Setup

### Connect to the server
```{r, include = FALSE}
library(tidyverse)
```

I use the keyring package to keep my credentials out of my R code. You can use the [great documentation available from RStudio](https://db.rstudio.com/best-practices/managing-credentials/#encrypt-credentials-with-keyring) to learn how do the same.

```{r}
user_name <-
  keyring::key_list("dbi_test")$username

con <- DBI::dbConnect(
      odbc::odbc(),
      Driver = "SQL Server",
      Server = "testsqlserver",
      UID = user_name,
      PWD = keyring::key_get("dbi_test", user_name)
   )
```

### Write some sample data 

Note that I set the temporary argument to TRUE so that the data is written to the tempdb on SQL server, which will result in it being deleted on disconnection.

This results in dplyr prefixing the table name with, "##."

SOURCE: https://db.rstudio.com/dplyr/#connecting-to-the-database 

```{r}
dplyr::copy_to(con, nycflights13::flights, "flights",
  temporary = TRUE
) 

dplyr::copy_to(con, nycflights13::airlines, "airlines",
  temporary = TRUE
) 

dplyr::copy_to(con, nycflights13::planes, "planes",
  temporary = TRUE
) 
```
## Option 1: Use dplyr syntax and let dbplyr handle the rest

### When I use this option

This is my default option. 

I do almost all of my analysis in R and this avoids fragmenting my work and thoughts across different tools.

### Examples

Example 1: filter rows, and retrieve selected columns
```{r}
dplyr::tbl(con, "##flights") %>%
  dplyr::filter(origin == "JFK", dest == "BQN" ) %>% 
  dplyr::select(flight, tailnum, dep_time, sched_dep_time,
                dep_delay) %>% 
  # I will remove the following line if exploring the data and
  # not interested in actually retrieving all records
  dplyr::collect() 
```

Example 2: join across tables and retrieve selected columns

```{r}
dplyr::tbl(con, "##flights") %>%
  dplyr::select(origin, dest, carrier, tailnum ) %>% 
  dplyr::left_join( #dplyr can join database tables 
    dplyr::tbl(con, "##airlines"), by = "carrier"
    ) %>% 
  mutate(airline = name) %>% 
  select(-carrier) %>% # only want the "human readable" airline name
  dplyr::collect()
```

Example 3: Summarize and count
```{r}
dplyr::tbl(con, "##flights") %>% 
  filter(!is.na(tailnum)) %>% 
  dplyr::left_join(tbl(con, "##planes"), by = "tailnum") %>% 
  dplyr::group_by(engine) %>% 
  dplyr::summarise(flights = n()) %>%
  dplyr::collect() %>% 
  dplyr::arrange(desc(flights))
```

Quite a few tailnum values in flights, are not present in planes, interesting!

## Option 2: Write SQL syntax and have dplyr and dbplyr run the query

### When I use this option

I use this option when I am reusing a *fairly short*, existing SQL query 
with minor modifications.

Example 1: Simple selection of records using SQL syntax
```{r}
query <- "SELECT  flight, tailnum, dep_time, origin, dest 
            FROM ##flights"

dplyr::tbl(con, dplyr::sql(query)) %>% 
  dplyr::collect()

```

Example 2: Use dplyr syntax to enhance a raw SQL query
```{r}
dplyr::tbl(con, dplyr::sql(query)) %>% #reuse query from previous example
  # add a filter state on top of the raw sql statement
  dplyr::filter(origin == "JFK", dest == "BQN" ) %>% 
  collect()
```

## Option 3: Store the SQL query in a text file and have dplyr and dbplyr run the query

### When I use this option

I use this approach under the following conditions:

1. I'm reusing existing SQL code or when collaborating with someone who will be writing new code in SQL
2. The SQL code is longer than a line or two

I prefer to, "modularize," my R code. Having an extremely long SQL statement 
in my R code doesn't abstract away the complexity of the SQL query. Putting the
query into it's own file helps achieve my desired level of abstraction. 

In conjunction with source control it makes tracking changes to the definition of a
data set simple.

**More importantly, it's a really useful way to collaborate with others who 
are comfortable with SQL but don't use R.** For example, I recently used this 
approach on a project involving aggregation of multiple data sets. 
Another team member focused on building out the data collection logic for 
some of the data sets in SQL. Once he had them built and validated he handed off
the query to me and I pasted it into a text file.

Step 1: Put your SQL code into a text file

Here is some example SQL code that might be in a file
```{sql, eval = FALSE}
SELECT f.year
        ,f.month
        ,f.day
        ,f.dep_time
        ,a.name
        ,p.manufacturer
        ,p.model
  FROM ##flights f
  LEFT JOIN ##airlines a ON f.carrier = a.carrier
  LEFT JOIN ##planes p ON f.tailnum = p.tailnum
  WHERE p.manufacturer IS NOT NULL 
```

Let's say that SQL code was stored in a text file called, `flights.sql`

Step 2: Use the SQL code in the file to retrieve data and execute the query.

```{r, eval = FALSE}
#reference: https://stackoverflow.com/a/46105261/7416441 
query <- readr::read_file("flights.sql") 

dplyr::tbl(con, dplyr::sql(query))
```

```{r, echo = FALSE}
query <- 
  "SELECT f.year
        ,f.month
        ,f.day
        ,f.dep_time
        ,a.name
        ,p.manufacturer
        ,p.model
  FROM ##flights f
  LEFT JOIN ##airlines a ON f.carrier = a.carrier
  LEFT JOIN ##planes p ON f.tailnum = p.tailnum
  WHERE p.manufacturer IS NOT NULL"

dplyr::tbl(con, dplyr::sql(query)) %>% 
  collect()
```
